A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-06-12 papers/water-SAFT/paper.tex
Added more papers to table (for organized citing) and improved pressure-density plot in paper.tex. Fixed an error in paper.bib. Tested everything on quipu.

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical,
  fu2005vapor-liquid-dft,felipe2001examination}, each of which
captures some of the qualitative behavior of water.  However, each of
these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A number of classical
density functionals have recently been produced that are based on
Statistical Associating Fluid Theory (SAFT)\cite{gloor2002saft,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, felipe2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

@@

Statistical Associating Fluid Theory (SAFT) is a set of models for
complex fluids in which hydrogren bonding plays a significant
role\cite{muller2001molecular}.  SAFT is used to accurately model the
equations of state of pure fluids and predict the equations of state
for mixtures over a wide range of temperatures and pressures.  SAFT is
based on Wertheim's first-order thermodynamic perturbation theory
(TPT1) \cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV} which allows it to
account for strong associative interations between molecules.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}. A majority of the empirical
parameters used in our functional are taken from the Clark\emph{et al}
SAFT functional which contains five empirical parameters and give the
correct pressure at liquid-vapor coexistince curve shown in Figure
\ref{fig:pressure-with-isotherms}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms-truncated}
\end{center}
\caption{The pressure versus density for various temperatures, including
experimental pressure data from NIST\cite{nistwater}. The dash-dotted lines
indicate the computationally calculated pressure and the open circles are 
NIST data points. The solid and dotted lines represent the theoretical
and experimental coexistence curves.}
\label{fig:pressure-with-isotherms}
\end{figure}

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, however, they also overestimate
the surface tension by a factor of about 1.5, and \fixme{most?} are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density approximation.

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact value
theorem and its limiting case of small solutes (Equation
\ref{contactvaluethm}). These properties are retained by the total
functional, provided the remaining terms are \emph{purely nonlocal}.

@@

The association functional we use is a modified version of Yu
and Wu.\cite{yu2002fmt-dft-inhomogeneous-associating} \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.} 
This functional includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$, but is
based on the SAFT-HS equation of state rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR}. SAFT-VR is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}, but to use it's
equation of state involves adding a correction term (see
Equation~\ref{eq:gSW}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-single-rod}
\end{center}
\caption{ Density profiles for a single rod of different radii. The dotted line
represents the saturated liquid density and the points represent the
expected contact density derived from the contact value theorem and
the free energy data
data plotted in Figure \ref{fig:energy-vs-diameter}.}
\label{fig:density-single-rod}
\end{figure}

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density profiles for
different radii rods which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. We use the
contact value theorem to derive the contact density at the water-rod
interface.  We start with the thermodyamic expression relating
pressure to the partial derivative of free energy with respect the
radius of a cylinder. Using the contact value theorem and sustituting in
the free energy gradient for the pressure, we can derive values for
the contact density from data used in Figure
\ref{fig:energy-vs-diameter}.  As the radius of the rods approch zero we
expect the contact density to approach the bulk density, which our
functional does.  This self consistency shows that our functional is
thermodynamically correct.

@@

To determine the critical separation, we start by consider a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for those
larger radius 
rods, the water on the sides of the 'stadium' configuration will bow inward 
between the rods and the density will reduce to vapor near the center
point where the rods are closest to each other.

Starting from the surface energy term, we can calculate the 
free energy per length, which is equal to the circumference multiplied 
by the surface tenson. The force per length is the derivative of this
with respect to the separation. 
We approximate the circumference of the stadium-shape by
\begin{equation}
C_{s} = 2\pi r +4r+2d
\end{equation}
and so the force per length is equal to $2\gamma$.

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use the aforementioned model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown graphically in 
Figure~\ref{fig:rods-energy-vs-distance} where there is a transition
betwen the sloped and flat portion of each plot. They match fairly well for the 
0.5~-~1.0~nm rods with differences between 1.3~-~5.5\% and we hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction
well (32\% difference).

@@

Walther \emph{et al.}\cite{walther2004hydrodynamic} studied the interactions between two
carbon nanotubes, which are geometrically similar to our hydrophobic rods. 
They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25~nm and separations ranging from about 0.3~nm to 1.5~nm.
They also find that a drying transition occurs, and find it to be at separations near
0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is larger than
what we find for hydrophobic rods of a similar size, as expected due to the carbon-carbon
and carbon-water interations.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ Free energy of hydradon 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/sphere-energy-vs-diameter}
\end{center}
\caption{ Free energy versus radius for a single hydrophobic sphere
  immersed in water. This should have an asymptote equal to the
  surface tension at room temperature, and it agrees well with the
  surface tension in Figure~\ref{fig:surface-tension}. Results from a
  simulation of SPC/E water~\cite{huang2001shs} are shown as circles.
  The horizonal lines show the experimental and SPC/E bulk surface
  tension for water at standard atmospheric temperature and
  pressure. }
\label{fig:sphere-energy-vs-diameter}
\end{figure}

@@

The classical density functional we developed by combining fundamental-measure 
theory with SAFT is able to produce reasonable results for hydrophobic interations.
We added the parameter $\lscale$ to the functional and fit it to the surface tension
at room temperature, which we find to be consistent with results for larger hydrophobic
rods.

>>>>

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,
  blas2001examination}, each of which
captures some of the qualitative behavior of water.  However, each of
these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A number of classical
density functionals have recently been produced that are based on
Statistical Associating Fluid Theory (SAFT)\cite{ 
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

@@

Statistical Associating Fluid Theory (SAFT) is a set of models for
complex fluids in which hydrogen bonding plays a significant
role\cite{muller2001molecular}.  SAFT is used to accurately model the
equations of state of pure fluids and predict the equations of state
for mixtures over a wide range of temperatures and pressures.  SAFT is
based on Wertheim's first-order thermodynamic perturbation theory
(TPT1) \cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV} which allows it to
account for strong associative interactions between molecules.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}. A majority of the empirical
parameters used in our functional are taken from the Clark \emph{et al}
paper which contains five empirical parameters and gives
correct pressures at the liquid-vapor coexistence curve shown in Figure
\ref{fig:pressure-with-isotherms}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms-truncated}
\end{center}
\caption{The pressure versus density for various temperatures, including
experimental pressure data from NIST\cite{nistwater}. The solid colored lines
indicate the computationally calculated pressure and the dotted
colored lines are NIST data points. The solid and dotted black lines
represent the theoretical and experimental coexistence curves.}
\label{fig:pressure-with-isotherms}
\end{figure}

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{clark2006developing, gloor2004accurate,
  gloor2007prediction, blas2001examination, gloor2002saft}.  Such
functionals have qualitatively predicted the dependence of surface
tension on temperature, however, they also overestimate
the surface tension by a factor of about 1.5, and most are unsuited
for studying systems that have density variations on a molecular
length scale due to the use of a local density
approximation\cite{gloor2002saft,clark2006developing, gloor2007prediction,
gloor2004accurate, gross2009density, kahl2008modified, blas2001examination}.

\begin{table}
\begin{tabular}{|c|c|c|c|c|}
\hline
paper & SAFT functional & water & LDA & tension \\
\hline
yuwu2002dft & X & & & \\
\hline
fuwu2005 & X & X & & X \\
\hline
kiselev2006 &  & X & X & X \\
\hline
clark2006developing & X & X & X & X \\
\hline
gloor2002 & X & X & X & X \\
\hline
gloor2004 & X & X & X & X \\
\hline
gloor2007 & X & X & X & X \\
\hline
gross2009 & X &  & X & X \\
\hline
Lischner2010 &  & X &  &  \\
\hline
Blas2001 & X & X & X & X \\
\hline
Kahl2008 & X &  & X & X \\
\hline
Muller2001 & X &  &  &  \\
\hline
Yang1994 &  & X &  & X \\
\hline
ding1987 &  & X & ? &  \\
\hline
next & ? & ? & ? & ? \\
\hline
\end{tabular}
\caption{Which papers do what}
\end{table}

@@

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional, which
incorporates effects of kinetic energy.  The ideal gas functional is
given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the
quantum concentration
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}.
\end{equation}
The ideal gas free energy functional satisfies the contact value
theorem and its limiting case of small solutes (Equations
\ref{eq:contact} and \ref{contactvaluethm}). These properties are
retained by the total functional, provided the remaining terms are
\emph{purely nonlocal}.

@@

The association functional we use is a modified version of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}. \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.}
This functional includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$, but is
based on the SAFT-HS equation of state rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR}. SAFT-VR is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}, but to use it's
equation of state involves adding a correction term (see
Equation~\ref{eq:gSW}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/density-single-rod}
\end{center}
\caption{ Density profiles for single rods of different radii. The dotted line
represents the saturated liquid density and the points represent the
expected contact density derived from the contact value theorem and
calculated free energy data.}
\label{fig:density-single-rod}
\end{figure}

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density profiles for
different radii rods which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. We use the
contact value theorem to derive the contact density at the water-rod
interface.  We start with the thermodynamic expression relating
pressure to the partial derivative of free energy with respect the
radius of a cylinder. Using the contact value theorem and substituting in
the free energy gradient for the pressure, we can derive values for
the contact density from data used in Figure
\ref{fig:energy-vs-diameter}.  As the radius of the rods approach zero we
expect the contact density to approach the bulk density, which our
functional does.  This self consistency shows that our functional is
thermodynamically correct.

@@

To determine the critical separation, we start by considering a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for those
larger radius 
rods, the water on the sides of the 'stadium' configuration will bow inward 
between the rods and the density will reduce to vapor near the center
point where the rods are closest to each other.

Starting from the surface energy term, we can calculate the 
free energy per length, which is equal to the circumference multiplied 
by the surface tension. The force per length is the derivative of this
with respect to the separation. 
We approximate the circumference of the stadium-shape by
\begin{equation}
C_{s} = 2\pi r +4r+2d
\end{equation}
and so the force per length is equal to $2\gamma$.

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use the aforementioned model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown graphically in 
Figure~\ref{fig:rods-energy-vs-distance} where there is a transition
between the sloped and flat portion of each plot. They match fairly well for the 
0.5~-~1.0~nm rods with differences between 1.3~-~5.5\% and we hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction
well (32\% difference).

@@

Walther \emph{et al.}\cite{walther2004hydrodynamic} studied the interactions between two
carbon nanotubes, which are geometrically similar to our hydrophobic rods. 
They use molecular dynamics with Lennard-Jones potentials for
nanotubes of diameter 1.25~nm and separations ranging from about 0.3~nm to 1.5~nm.
They also find that a drying transition occurs, and find it to be at separations near
0.9~-~1.0~nm\cite{walther2004hydrodynamic}. This is larger than
what we find for hydrophobic rods of a similar size, as expected due to the carbon-carbon
and carbon-water interactions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ Free energy of hydration 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/sphere-energy-vs-diameter}
\end{center}
\caption{ Free energy versus radius for a single hydrophobic sphere
  immersed in water. This should have an asymptote equal to the
  surface tension at room temperature, and it agrees well with the
  surface tension in Figure~\ref{fig:surface-tension}. Results from a
  simulation of SPC/E water~\cite{huang2001shs} are shown as circles.
  The horizontal lines show the experimental and SPC/E bulk surface
  tension for water at standard atmospheric temperature and
  pressure. }
\label{fig:sphere-energy-vs-diameter}
\end{figure}

@@

The classical density functional we developed by combining fundamental-measure 
theory with SAFT is able to produce reasonable results for hydrophobic interactions.
We added the parameter $\lscale$ to the functional and fit it to the surface tension
at room temperature, which we find to be consistent with results for larger hydrophobic
rods.
